---
title: "Dataset Cleanup and Conversion to external_gene_name"
output: html_document
---

Set working directory
```{r set directory}
knitr::opts_knit$set(root.dir = '~/bioinformatics/InVivo/')
knitr::opts_chunk$set(results = 'hide')
```

Load libraries
```{r libraries}
library(tidyverse)
library(readxl)
```

```{bash}
pwd
```

## Read in Datasets
*adjust file paths as necessary*
```{r GSE134196}
setwd('~/bioinformatics/InVivo/')
GSE.134196 <- read.csv("Selected_GEO_datasets/GSE134196_counts.L4440.Y37A1B_5.csv")
GSE.134196
```
<!-- wormbase_gene -->
<!-- 46,739 -->

```{r GSE156037}
setwd('~/bioinformatics/InVivo/')
GSE.156037 <- read.csv("Selected_GEO_datasets/GSE156037_raw_counts.csv")
GSE.156037
```
<!-- wormbase_gene -->
<!-- 46,093 -->

```{r GSE41205}
setwd('~/bioinformatics/InVivo/')
GSE.41205_N2genes <- read.delim("Selected_GEO_datasets/GSE41205_Paul_2012_MAPK/GSM1010460_N2genes.txt", 
                                sep = "\t")
GSE.41205_KU25genes <- read.delim("Selected_GEO_datasets/GSE41205_Paul_2012_MAPK/GSM1010461_KU25genes.txt", 
                                  sep = "\t")
GSE.41205_N2genes
GSE.41205_KU25genes
```
<!-- external_gene_name -->
<!-- 9,117 -->
<!-- 8,820 -->

```{r GSE46257}
setwd('~/bioinformatics/InVivo/')
GSE.46257 <- read.csv("Selected_GEO_datasets/GSE46257_Schmeisser_2013_arsenite/GSE46257_ce_Nr35-38_counts_rpkm.csv")
GSE.46257
```
<!-- external_gene_name -->
<!-- 19,912 genes -->

```{r GSE63528}
setwd('~/bioinformatics/InVivo/')
GSE.63528_1 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551786_Sample_1_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_2 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551787_Sample_2_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_3 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551788_Sample_3_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_4 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551789_Sample_4_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_5 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551790_Sample_5_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_6 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551791_Sample_6_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_7 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551792_Sample_7_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_8 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551793_Sample_8_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_9 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551794_Sample_9_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_10 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551795_Sample_10_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_11 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551796_Sample_11_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_12 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551797_Sample_12_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_13 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551798_Sample_13_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_14 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551799_Sample_14_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_15 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551800_Sample_15_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_16 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551801_Sample_16_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_17 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551802_Sample_17_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_18 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551803_Sample_18_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_19 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551804_Sample_19_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_20 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551805_Sample_20_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_21 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551806_Sample_21_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_22 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551807_Sample_22_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_23 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551808_Sample_23_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_24 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551809_Sample_24_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_25 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551810_Sample_25_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_26 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551811_Sample_26_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_27 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551812_Sample_27_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_28 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551813_Sample_28_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_29 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551814_Sample_29_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_30 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551815_Sample_30_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_31 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551816_Sample_31_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_32 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551817_Sample_32_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_33 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551818_Sample_33_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_34 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551819_Sample_34_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_35 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551820_Sample_35_processed_data.txt", 
                          sep = "\t", row.names = 1)
GSE.63528_36 <- read.delim("Selected_GEO_datasets/GSE63528_Rangaraju_2016/GSM1551821_Sample_36_processed_data.txt", 
                          sep = "\t", row.names = 1)

GSE.63528_1
GSE.63528_2
GSE.63528_3
GSE.63528_4
GSE.63528_5
GSE.63528_6
GSE.63528_7
GSE.63528_8
GSE.63528_9
GSE.63528_10
GSE.63528_11
GSE.63528_12
GSE.63528_13
GSE.63528_14
GSE.63528_15
GSE.63528_16
GSE.63528_17
GSE.63528_18
GSE.63528_19
GSE.63528_20
GSE.63528_21
GSE.63528_22
GSE.63528_23
GSE.63528_24
GSE.63528_25
GSE.63528_26
GSE.63528_27
GSE.63528_28
GSE.63528_29
GSE.63528_30
GSE.63528_31
GSE.63528_32
GSE.63528_33
GSE.63528_34
GSE.63528_35
GSE.63528_36
```
<!-- external_gene_name -->
<!-- 20,537 genes -->

<!-- WormSPELL sets -->
```{r GSE49662}
setwd('~/bioinformatics/InVivo/')
GSE.49662 <- read.delim("Selected_GEO_datasets/WormSPELL/GSE49662_ce_Nr101-109_counts.txt", 
                          sep = "\t")
GSE.49662
```
<!-- external_gene_name -->
<!-- 45,461 -->


```{r GSE55988}
setwd('~/bioinformatics/InVivo/')
GSE.55988 <- read.delim("Selected_GEO_datasets/WormSPELL/GSE55988_htseq-count-table.txt", 
                          sep = "\t")
GSE.55988  
```
<!-- external_gene_name -->
<!-- 45,836 -->

```{r GSE46051 xls}
setwd('~/bioinformatics/InVivo/')
GSE.46051 <- read_xls("Selected_GEO_datasets/WormSPELL/GSE46051_ce_22samples_DMSO_Rotenone_counts_and_RPKM.xls")

GSE.46051
```
<!-- external_gene_name -->
<!-- 45,461 -->


```{r GSE54853 xls}
setwd('~/bioinformatics/InVivo/')
GSE.54853 <- read_xls("Selected_GEO_datasets/WormSPELL/GSE54853_Nr01-12_25-36_counts_ce_mm.xls")
GSE.54853
```
<!-- external_gene_name -->
<!-- 19,911 -->

<!-- InVivo data -->
```{r invivo data}
setwd('~/bioinformatics/InVivo/')
invivo <- read.delim("../group-projects-invivo-fall-project/BGMP_IVB_2020_dataset/merged_counts.tsv",
                     sep = "\t")
# remove last 5 lines that don't contain genecounts
invivo <- invivo[1:(nrow(invivo)-5),]
```
<!-- wormbase_gene -->
<!-- 46,909 -->



## Cleanup for WGCNA
Select only relevant columns
**be careful to run this chunk only once, otherwise the row numbers will replace the gene names*

```{r}
ID_df <- read.delim("~/bioinformatics/group-projects-invivo-fall-project/Cleaned_Datasets_transcript_stableID/celegans_cleaned_mart_ensembl.102.tsv", sep = "\t")
ID_df
```



```{r GSE46257 cleanup}
GSE.46257_converted <- merge(GSE.46257, ID_df, by.x = "X", by.y = "Gene.name", all.x = FALSE, all.y = FALSE)
GSE.46257_converted <- GSE.46257_converted %>%
  select(Transcript.stable.ID, ends_with("counts")) %>%
  rename("transcript_stable_ID" = Transcript.stable.ID)
GSE.46257_converted

nrow(GSE.46257)
nrow(GSE.46257_converted)
nrow(GSE.46257_converted)/nrow(GSE.46257)
```

```{r GSE63528 cleanup}
# select only the counts columns
GSE.63528_1 <- GSE.63528_1 %>% select(starts_with("Count"))
GSE.63528_2 <- GSE.63528_2 %>% select(starts_with("Count"))
GSE.63528_3 <- GSE.63528_3 %>% select(starts_with("Count"))
GSE.63528_4 <- GSE.63528_4 %>% select(starts_with("Count"))
GSE.63528_5 <- GSE.63528_5 %>% select(starts_with("Count"))
GSE.63528_6 <- GSE.63528_6 %>% select(starts_with("Count"))
GSE.63528_7 <- GSE.63528_7 %>% select(starts_with("Count"))
GSE.63528_8 <- GSE.63528_8 %>% select(starts_with("Count"))
GSE.63528_9 <- GSE.63528_9 %>% select(starts_with("Count"))
GSE.63528_10 <- GSE.63528_10 %>% select(starts_with("Count"))
GSE.63528_11 <- GSE.63528_11 %>% select(starts_with("Count"))
GSE.63528_12 <- GSE.63528_12 %>% select(starts_with("Count"))
GSE.63528_13 <- GSE.63528_13 %>% select(starts_with("Count"))
GSE.63528_14 <- GSE.63528_14 %>% select(starts_with("Count"))
GSE.63528_15 <- GSE.63528_15 %>% select(starts_with("Count"))
GSE.63528_16 <- GSE.63528_16 %>% select(starts_with("Count"))
GSE.63528_17 <- GSE.63528_17 %>% select(starts_with("Count"))
GSE.63528_18 <- GSE.63528_18 %>% select(starts_with("Count"))
GSE.63528_19 <- GSE.63528_19 %>% select(starts_with("Count"))
GSE.63528_20 <- GSE.63528_20 %>% select(starts_with("Count"))
GSE.63528_21 <- GSE.63528_21 %>% select(starts_with("Count"))
GSE.63528_22 <- GSE.63528_22 %>% select(starts_with("Count"))
GSE.63528_23 <- GSE.63528_23 %>% select(starts_with("Count"))
GSE.63528_24 <- GSE.63528_24 %>% select(starts_with("Count"))
GSE.63528_25 <- GSE.63528_25 %>% select(starts_with("Count"))
GSE.63528_26 <- GSE.63528_26 %>% select(starts_with("Count"))
GSE.63528_27 <- GSE.63528_27 %>% select(starts_with("Count"))
GSE.63528_28 <- GSE.63528_28 %>% select(starts_with("Count"))
GSE.63528_29 <- GSE.63528_29 %>% select(starts_with("Count"))
GSE.63528_30 <- GSE.63528_30 %>% select(starts_with("Count"))
GSE.63528_31 <- GSE.63528_31 %>% select(starts_with("Count"))
GSE.63528_32 <- GSE.63528_32 %>% select(starts_with("Count"))
GSE.63528_33 <- GSE.63528_33 %>% select(starts_with("Count"))
GSE.63528_34 <- GSE.63528_34 %>% select(starts_with("Count"))
GSE.63528_35 <- GSE.63528_35 %>% select(starts_with("Count"))
GSE.63528_36 <- GSE.63528_36 %>% select(starts_with("Count"))

# group all samples into a list
GSE.63528_list <- c(
  GSE.63528_1, GSE.63528_2, GSE.63528_3, GSE.63528_4, GSE.63528_5,
  GSE.63528_6, GSE.63528_7, GSE.63528_8, GSE.63528_9, GSE.63528_10,
  GSE.63528_11, GSE.63528_12, GSE.63528_13, GSE.63528_14, GSE.63528_15,
  GSE.63528_16, GSE.63528_17, GSE.63528_18, GSE.63528_19, GSE.63528_20,
  GSE.63528_21, GSE.63528_22, GSE.63528_23, GSE.63528_24, GSE.63528_25,
  GSE.63528_26, GSE.63528_27, GSE.63528_28, GSE.63528_29, GSE.63528_30,
  GSE.63528_31, GSE.63528_32, GSE.63528_33, GSE.63528_34, GSE.63528_35,
  GSE.63528_36
)


# bind columns and convert tibble to df, re-add geneID
GSE.63528_merged <- as.data.frame(bind_cols(GSE.63528_list))
rownames(GSE.63528_merged) <- rownames(GSE.63528_1)
GSE.63528_merged <- rownames_to_column(GSE.63528_merged, var = "transcript_stable_ID")
GSE.63528_merged
```

```{r GSE49662 cleanup}
# drop extra gene ID column and rename kept one to transcript_stable_ID
GSE.49662 <- GSE.49662 %>%
  select("ensembl_gene_id", starts_with("Nr"))  %>%
  rename("transcript_stable_ID" = ensembl_gene_id)
GSE.49662
```

```{r GSE55988 cleanup}
# rename gene ID column
GSE.55988 <- GSE.55988 %>% rename("transcript_stable_ID" = GeneID)
GSE.55988
```

```{r GSE46051 cleanup}
# drop extra gene ID columns and rename kept one
GSE.46051 <- GSE.46051 %>% 
  select("gene", starts_with("JA")) %>% 
  rename("transcript_stable_ID" = gene)
GSE.46051
```

```{r GSE54853 cleanup}
# rename gene ID column
GSE.54853_converted <- merge(GSE.54853, ID_df, by.x = "gene", by.y = "Gene.name", all.x = FALSE, all.y = FALSE)
GSE.54853_converted <- GSE.54853_converted %>%
  select(Transcript.stable.ID, starts_with("Nr")) %>%
  rename("transcript_stable_ID" = Transcript.stable.ID)
GSE.54853_converted

nrow(GSE.54853)
nrow(GSE.54853_converted)
nrow(GSE.54853_converted)/nrow(GSE.54853)
```

```{r GSE41205 cleanup}
# GSE41205
# select only counts columns, rename gene ID column 
GSE.41205_N2genes <- GSE.41205_N2genes %>%
  select("GeneID", starts_with("Uniq_reads")) %>%
  rename("transcript_stable_ID" = GeneID)
GSE.41205_KU25genes <- GSE.41205_KU25genes %>%
  select("GeneID", starts_with("Uniq_reads")) %>%
  rename("transcript_stable_ID" = GeneID)

# need to use merge function because different number of rows
GSE.41205_N2genes
GSE.41205_KU25genes
GSE.41205_merged <- merge(GSE.41205_N2genes, GSE.41205_KU25genes, by="transcript_stable_ID")
GSE.41205_merged
```


## Convert Wormbase gene IDs to external_gene_name
Convert IDs
```{r geneID to transcriptID GSE.134196}
GSE.134196_converted <- merge(GSE.134196, ID_df, by.x = "X", by.y = "Gene.stable.ID", all.x = FALSE, all.y = FALSE)
GSE.134196_converted <- GSE.134196_converted %>%
  select(Transcript.stable.ID, starts_with("L"), starts_with("Y")) %>%
  rename("transcript_stable_ID" = Transcript.stable.ID)

GSE.134196_converted

# check retention
nrow(GSE.134196)
nrow(GSE.134196_converted)
nrow(GSE.134196_converted) / nrow(GSE.134196)
```
<!-- original number of genes: 46,739 -->
<!-- number of genes after conversion to external gene name: 46,571 -->
<!-- proportion retained: 0.9964056 -->

```{r geneID to transcriptID GSE156037}
GSE.156037_converted <- merge(GSE.156037, ID_df, by.x = "rowname", by.y = "Gene.stable.ID", all.x = FALSE, all.y = FALSE)
GSE.156037_converted <- GSE.156037_converted %>%
  select(Transcript.stable.ID, starts_with("S")) %>%
  rename("transcript_stable_ID" = Transcript.stable.ID)

GSE.156037_converted

nrow(GSE.156037)
nrow(GSE.156037_converted)
nrow(GSE.156037_converted) / nrow(GSE.156037)
```
<!-- original number of genes: 46,093 -->
<!-- number of genes after conversion to external gene name: 45,928 -->
<!-- proportion retained: 0.9964203 -->

```{r geneID to transcriptID invivo data}
invivo_converted <- merge(invivo, ID_df, by.x = "feature", by.y = "Gene.stable.ID", all.x = FALSE, all.y = FALSE)
invivo_converted <- invivo_converted %>%
  select(Transcript.stable.ID, starts_with("X")) %>%
  rename("transcript_stable_ID" = Transcript.stable.ID)

invivo_converted

nrow(GSE.156037)
nrow(GSE.156037_converted)
nrow(GSE.156037_converted) / nrow(GSE.156037)
```
<!-- original number of genes: 46,909 -->
<!-- number of genes after conversion to external gene name: 46,904 -->
<!-- proportion retained: 0.9998934 -->

### Check Samples Before Writing Out
```{r check before write out}
GSE.46257_converted
GSE.49662
GSE.55988
GSE.46051
GSE.54853_converted
GSE.41205_merged
GSE.63528_merged

GSE.134196_converted
GSE.156037_converted
invivo_converted
```


### Write Out to CSV Files
<!-- GSE.46257 -- mix of lowercase gene names and uppercase gene-synonym-looking IDs -->
<!-- GSE.49662 -->
<!-- GSE.55988 -->
<!-- GSE.46051 -->
<!-- GSE.54853 -- mix of lowercase gene names and uppercase gene-synonym-looking IDs -->
<!-- GSE.41205_merged -->
<!-- GSE.63528_merged -->
<!-- GSE.134196 -- were WB genes, converted to external gene name -->
<!-- GSE.156037 -- were WB genes, converted to external gene name -->
<!-- invivo -- were WB genes, converted to external gene name -->
*change directory as needed*
```{r write out files}
# set directory for where files should be written to
setwd("~/bioinformatics/group-projects-invivo-fall-project/Cleaned_Datasets_transcript_stableID/")

write_csv(GSE.46257_converted, "GSE46257_cleaned_transcrID.csv")
write_csv(GSE.49662, "GSE49662_cleaned_transcrID.csv")
write_csv(GSE.55988, "GSE55988_cleaned_transcrID.csv")
write_csv(GSE.46051, "GSE46051_cleaned_transcrID.csv")
write_csv(GSE.54853_converted, "GSE54853_cleaned_transcrID.csv")
write_csv(GSE.41205_merged, "GSE41205_cleaned_transcrID.csv")
write_csv(GSE.63528_merged, "GSE63528_cleaned_transcrID.csv")
write_csv(GSE.134196_converted, "GSE134196_cleaned_transcrID.csv")
write_csv(GSE.156037_converted, "GSE156037_cleaned_transcrID.csv")
write_csv(invivo_converted, "invivo_cleaned_transcrID.csv")
```





